---
title: "STAT 434 Project Modeling"
author: "Emma Barton and Ryan Vosbigian"
date: "5/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Types of models to use:

Classification models only (could conver classes to numeric 1 thru 4, but makes our prediction undesirable, plus not equal jumps between classes)

  1. Logistic Regression (pairwise/one-versus-one and one-versus-all)?
  2. Linear Discriminant Analysis
  3. Quadratic Discriminant Analysis
  4. K-Nearest Neighbor
  5. Classification Tree
  6. Bagged Trees
  7. Random Forests
  8. Boosted Forests
  9. Neural Networks
  10. Support Vector Machines
  
Additional Considerations for Modeling

  1. PCA of all diet and foraging variables
    a. All diet variables are highly correlated, because they each sum up to 100. Thus, using PCA could really simplify the models considerably
    b. Additionally, if PCA is performed on diet variables then the diet variables should not be scaled because they already are by nature of the data
  2. Changing sensitivity of classifications so it is harder to be classified as Least Concern
    a. ROC Curves for each model
  
# Set Up

```{r load packages}
library(tidyverse)

```

```{r set pathway}

## Local Pathways - copy/paste your local pathway to the github repository

# Emma
pathway = "/Users/emmabarton/Desktop/spring_2020/STAT434/project/"

# Ryan
#pathway = "~/GitHub/STAT-434---not-the-dodo-/"


```

Loading data and combining exctinct species and critically endangered species. 

```{r load data}

mammal_df = read.csv("cleaned_mammal_df.csv")
bird_df = read.csv("cleaned_bird_df_updated_5_29.csv")

#combine "Extinct" & "Extinct in the Wild" & "Critically Endangered" into "Endangered"
mammal_df$redlistCategory = as.character(car::recode(mammal_df$redlistCategory,recodes = "'Extinct'='Endangered';'Extinct in the Wild'='Endangered';'Critically Endangered'='Endangered'"))

bird_df$redlistCategory = as.character(car::recode(bird_df$redlistCategory,recodes = "'Extinct'='Endangered';'Extinct in the Wild'='Endangered';'Critically Endangered'='Endangered'"))

#converting to as.factor

mammal_df$Activity.Nocturnal = as.factor(mammal_df$Activity.Nocturnal)
mammal_df$Activity.Crepuscular = as.factor(mammal_df$Activity.Crepuscular)
mammal_df$Activity.Diurnal = as.factor(mammal_df$Activity.Diurnal)

bird_df$Nocturnal = as.factor(bird_df$Nocturnal)
```

# Preparing Data

```{r}
#moving data deficient mammals to different data set
data_deficient_mammals_df = mammal_df[mammal_df$redlistCategory == "Data Deficient",]
#removing data deficient mammals
mammal_df = mammal_df[mammal_df$redlistCategory != "Data Deficient",]

#moving data def birds to different dataset
data_deficient_birds_df = bird_df[bird_df$redlistCategory == "Data Deficient",]
#removing data deficient birds
bird_df = bird_df[bird_df$redlistCategory != "Data Deficient",]

#creating indices for training sets

set.seed(10)

train_mammal = sample(1:nrow(mammal_df),size = round(nrow(mammal_df)/2))
train_bird = sample(1:nrow(bird_df),size = round(nrow(bird_df)/2))

# creating training and test sets

mammal.train.df = mammal_df[train_mammal,]
mammal.test.df = mammal_df[-train_mammal,]

bird.train.df = bird_df[train_bird,]
bird.test.df = bird_df[-train_bird,]

table(bird.train.df$redlistCategory)

```


## Testing PCA

Creating data frames with Principle Components as columns

```{r}


pcbirddiet = prcomp(bird_df[,c(12:21)],scale=FALSE)

pcbirdforage = prcomp(bird_df[,c(26:32)],scale=FALSE)

pcmammal = prcomp(mammal_df[,c(4:13)],scale = FALSE)

pcbird = prcomp(bird_df[,c(12:21,26:32)],scale=FALSE)

summary(pcmammal)


plot(cumsum(pcbirddiet$sdev^2/sum(pcbirddiet$sdev^2)), xlab="Principal  Component", ylab="Cumulative  Proportion  of Variance  Explained", ylim=c(0,1),type="b",main = "Bird Diet")

plot(cumsum(pcmammal$sdev^2/sum(pcmammal$sdev^2)), xlab="Principal  Component", ylab="Cumulative  Proportion  of Variance  Explained", ylim=c(0,1),type="b",main = "Mammal Diet")


plot(cumsum(pcbirdforage$sdev^2/sum(pcbirdforage$sdev^2)), xlab="Principal  Component", ylab="Cumulative  Proportion  of Variance  Explained", ylim=c(0,1),type="b",main = "Bird Foraging")

plot(cumsum(pcbird$sdev^2/sum(pcbird$sdev^2)), xlab="Principal  Component", ylab="Cumulative  Proportion  of Variance  Explained", ylim=c(0,1),type="b",main = "Bird Foraging and Diet")


colnames(pcbirddiet$x) = paste("Diet",colnames(pcbirddiet$x),sep = "")

colnames(pcmammal$x) = paste("Diet",colnames(pcmammal$x),sep = "")

colnames(pcbirdforage$x) = paste("For",colnames(pcbirdforage$x),sep = "")

colnames(pcbird$x) = paste("DietFor",colnames(pcbird$x),sep = "")

pc_bird_df = cbind(bird_df,
                   cbind(pcbirddiet$x,
                         cbind(pcbirdforage$x,pcbird$x)))

pc_mammal_df = cbind(mammal_df,pcmammal$x)


pc.mammal.train.df = pc_mammal_df[train_mammal,]
pc.mammal.test.df = pc_mammal_df[-train_mammal,]

pc.bird.train.df = pc_bird_df[train_bird,]
pc.bird.test.df = pc_bird_df[-train_bird,]

```


# 1: Logistic Regression
  
# 2: Linear Discriminant Analysis

Starting by using a bunch of variables and 


```{r}

library(MASS)


bird_lda1 = lda(redlistCategory∼Diet.Inv + Diet.Vend + Diet.Vect + Diet.Vfish + Diet.Vunk + Diet.Scav + Diet.Fruit + Diet.Nect + Diet.Seed + Diet.PlantO + Diet.5Cat + ForStrat.watbelowsurf + ForStrat.wataroundsurf + ForStrat.aerial +  PelagicSpecialist + Nocturnal + BodyMass.Value,data=bird.train.df)

bird_lda1_pred=predict(bird_lda1, bird.test.df)

table(preds = bird_lda1_pred$class,truth = bird.test.df$redlistCategory)

mean(bird_lda_pred$class!=bird.test.df$redlistCategory)
#error rate 0.2279793
colnames(pc.bird.train.df)
#Using first 11 PCs for Diet and Foraging PCs, because that is where the elbow is in the graph above
bird_lda_pca1 = lda(redlistCategory∼ DietForPC1 + DietForPC2 + DietForPC3 + DietForPC4 + DietForPC5 + DietForPC6 + DietForPC7 + DietForPC8 + DietForPC9 + DietForPC10 +DietForPC11 + Diet.5Cat +  PelagicSpecialist + Nocturnal + scale(BodyMass.Value),data=pc.bird.train.df)

bird_lda_pca1_pred=predict(bird_lda_pca1, pc.bird.test.df)

table(preds = bird_lda_pca1_pred$class,truth = bird.test.df$redlistCategory)

mean(bird_lda_pca1_pred$class!=bird.test.df$redlistCategory)
#error rate 0.2232124

```
# 3: Quadratic Discriminant Analysis

Getting errors because I have way to many predictors in model.

```{r}
bird_qda1 = qda(redlistCategory∼Diet.Inv + Diet.Vend + Diet.Vect + Diet.Vfish + Diet.Vunk + Diet.Scav + Diet.Fruit + Diet.Nect + Diet.Seed + Diet.PlantO + ForStrat.watbelowsurf + ForStrat.wataroundsurf + ForStrat.aerial +  PelagicSpecialist + Nocturnal + scale(BodyMass.Value),data=bird.train.df)

bird_qda1_pred=predict(bird_qda1, bird.test.df)

table(preds = bird_qda1_pred$class,truth = bird.test.df$redlistCategory)

mean(bird_qda_pred$class!=bird.test.df$redlistCategory)


bird_qda_pca1 = qda(redlistCategory∼ DietForPC1 + DietForPC2 + DietForPC3 + DietForPC4 + DietForPC5 + DietForPC6 + DietForPC7 + DietForPC8 + DietForPC9 + DietForPC10 +DietForPC11 +  PelagicSpecialist + Nocturnal + scale(BodyMass.Value),data=pc.bird.train.df)

bird_qda_pca1_pred=predict(bird_qda_pca1, pc.bird.test.df)

table(preds = bird_qda_pca1_pred$class,truth = bird.test.df$redlistCategory)

mean(bird_qda_pca1_pred$class!=bird.test.df$redlistCategory)
#test error: 0.2945078



```

# 4: K-Nearest Neighbor

```{r}



```



# 5: Classification Tree



# 6: Bagged Trees

# 7: Random Forests

# 8: Boosted Forests

# 9: Neural Networks

# 10: Support Vector Machines





```{r}

```




