---
title: "Data Cleaning"
author: "Ryan Vosbigian & Emma Barton"
date: "5/21/2020"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

# Set Up

```{r load packages}

library(tidyverse)
library(Hmisc)

```

```{r set pathway}

## Local Pathways - copy/paste your local pathway to the github repository

# Emma
pathway = "/Users/emmabarton/Desktop/spring_2020/STAT434/project/"

# Ryan
#pathway = "~/GitHub/"

```

## Loading Datasets

```{r raw data}

redlist_df = read.csv(paste0(pathway, "STAT-434---not-the-dodo-/simple_summary_IUCN_birds_mammals.csv"))

elton_mammal_df = read.csv(paste0(pathway, "STAT-434---not-the-dodo-/EltonTraits/MamFuncDat.txt"),sep = "\t")

elton_bird_df = read.csv(paste0(pathway, "STAT-434---not-the-dodo-/EltonTraits/BirdFuncDat.txt"),sep = "\t")

```

```{r change raw data types}

redlist_df$scientificName = as.character(redlist_df$scientificName)
elton_mammal_df$scientificName = as.character(elton_mammal_df$Scientific)
elton_bird_df$scientificName = as.character(elton_bird_df$Scientific)

```

```{r join elton and redlist}

# Filter mammals and birds from redlist species
redlist_mammals_df = redlist_df[redlist_df$className=="MAMMALIA",]
redlist_birds_df = redlist_df[redlist_df$className=="AVES",]

# Join Elton traits and redlist species for mammals
mammal_df = left_join(x = elton_mammal_df, y = redlist_mammals_df, 
                      by="scientificName")

# Join Elton traits and redlist species for birds
bird_df = left_join(x = elton_bird_df, y = redlist_birds_df,  
                    by="scientificName")

# Write data
write.csv(mammal_df, "uncleaned_mammal_df.csv")
write.csv(bird_df, "uncleaned_bird_df.csv")

```

```{r uncleaned data}

mammal_df = read.csv("uncleaned_mammal_df.csv")
bird_df = read.csv("uncleaned_bird_df.csv")

```

## Data Cleaning

```{r data dimensions}

dim(mammal_df)
dim(bird_df)

```

```{r missing redlist species}

table(is.na(mammal_df$Diet.Vect))
nrow(mammal_df)
nrow(elton_mammal_df)
nrow(redlist_mammals_df)

# Check redlist mammals that do not have Elton traits
redlist_mammals_df %>% 
  filter(!(scientificName %in% elton_mammal_df$scientificName)) %>% 
  nrow()

# Check redlist birds that do not have Elton traits
redlist_birds_df %>% 
  filter(!(scientificName %in% elton_bird_df$scientificName)) %>% 
  nrow()

```

```{r select columns}

# Removed unnecessary or mostly empty columns
mammal_df = mammal_df %>% 
  select(-X, -MSW3_ID, -infraType, -infraName, -infraAuthority, 
         -redlistCriteria, -criteriaVersion)

bird_df = bird_df %>% 
  select(-X, -infraType, -infraName, -infraAuthority, 
         -redlistCriteria, -criteriaVersion)

```

```{r missing redlist status}

mammal_df %>% 
  filter(is.na(redlistCategory)) %>% 
  nrow()

bird_df %>% 
  filter(is.na(redlistCategory)) %>% 
  nrow()

```

```{r filter out missing redlist status}

# Create prediction/test datasets 
mammal_pred_df = mammal_df %>% 
  filter(is.na(redlistCategory))

bird_pred_df = bird_df %>% 
  filter(is.na(redlistCategory))

# Filter out rows with missing redlist category to make training dataset
mammal_df = mammal_df %>% 
  filter(!is.na(redlistCategory))

bird_df = bird_df %>% 
  filter(!is.na(redlistCategory))

```

```{r write cleaned data}

write.csv(mammal_df, "cleaned_mammal_df.csv")
write.csv(bird_df, "cleaned_bird_df.csv")

write.csv(mammal_pred_df, "mammal_pred_df.csv")
write.csv(bird_pred_df, "bird_pred_df.csv")

```

Not having much luck matching these 500 mammal taxons. maybe just drop them??
```{r matching extra mammal species}

# matching MAMMAL species that didnt have a perfect match between the IUCN and EltonTraits databases
mammal_pred_df = read.csv("mammal_pred_df.csv")

syn_df = read.csv("asm-species-2020-05-27_containssynonymsfromeltontraits.csv")

#removing the empty columns from the missing IUCN join
mammal_pred_df = mammal_pred_df[,1:27]

#reformatting the deprecated.scientific name column from: {"herpestes javanica":":$2009"}

#chopping off the first two characters

syn_df$Deprecated.Scientific = gsub('^."([^"]*[^"]*)".*$', "\\1",syn_df$Deprecated.Scientific)

#capitalizing
syn_df$Deprecated.Scientific = capitalize(syn_df$Deprecated.Scientific)

mammal_pred_df %>% 
  filter(!(scientificName %in% syn_df$Deprecated.Scientific)) %>% 
  unique()

mammal_pred_df$scientificName = as.character(mammal_pred_df$scientificName)
syn_df$Deprecated.Scientific = as.character(syn_df$Deprecated.Scientific)
syn_df$Canonical.Sciname = as.character(syn_df$Canonical.Sciname)

for (i in 1:nrow(mammal_pred_df)){
  sci_index = match(mammal_pred_df$scientificName[i],syn_df$Deprecated.Scientific)
  if (is.na(sci_index)==FALSE) {
    mammal_pred_df$scientificName[i] = syn_df$Canonical.Sciname[sci_index]
  }else {
    mammal_pred_df$scientificName[i] = NA
  }
}

```

TaxonomicChanges_BirdLife_Avibase.csv was scrapped from avibase.bs-eoc.org
https://avibase.bsc-eoc.org/compare.jsp?source1=birdlife&version1=BIRDLIFE03&source2=birdlife&version2=BIRDLIFE12&continent=&reg_type=9

It is a table with that matches the Birdlife checklist version 3 (EltonTraits uses) with the HBW Birdlife Taxonomic Checklist v4 (which IUCN uses). Using this table to change names to the new taxonomic system to change the taxonomic system used in EltonTraits to the more updated v4. 

http://datazone.birdlife.org/species/taxonomy
^birdlife taxonomy

Legend at bottom of table
Taxon removed. This color indicates that the taxonomic concept is only present in the first checklist. The species was either removed entirely from the list, or may have been replaced by another taxon concept (eg, following a species split or a species lump).
 	Taxon added. This color indicates that the taxonomic concept is only present in the second checklist. The species was either entirely new to the checklist, or it may replace another concept (eg, following a species split or a species lump).
 	Name changed. This color indicates that the taxonomic concept is the same between the 2 versions (ie, that the names refers to the same population), but that the scientific, common or family name has been modified. 
 	
```{r matching extra bird species}

# matching BIRDS that didnt have a perfect match between the IUCN and EltonTraits databases
mammal_pred_df = read.csv("mammal_pred_df.csv")

bird_pred_df = read.csv("bird_pred_df.csv")

avibase_df = read.csv("TaxonomicChanges_Birdlife_Avibase.csv")
avibase_df = avibase_df[,2:ncol(avibase_df)] #chopping off the 
avibase_ioc_df = read.csv("TaxonomicChanges_IOC27_HBWBirdlife4.csv")
avibase_ioc_df$Scientific.name = avibase_ioc_df$'ï..Scientific.name'

#matching taxonomy from Birdlife v3 to HBW v4, filtering out ones that don't match (the ones using taxonomy IOC 2.7 don't match)
bird_pred_df %>% 
  filter((!scientificName %in% avibase_df$Scientific.name)) %>% 
  unique()

#looking at taxons that don't match within IOC2.7 or birdlife v3
bird_pred_df %>% 
  filter((!scientificName %in% avibase_df$Scientific.name)) %>% 
  filter((!scientificName %in% avibase_ioc_df$Scientific.name)) %>%
  unique()

#changing to character to avoid errors
bird_pred_df$scientificName = as.character(bird_pred_df$scientificName)
avibase_df$Scientific.name = as.character(avibase_df$Scientific.name)
avibase_df$Scientific.name2 = as.character(avibase_df$Scientific.name2)
avibase_ioc_df$Scientific.name = as.character(avibase_ioc_df$Scientific.name)
avibase_ioc_df$Scientific.name2 = as.character(avibase_ioc_df$Scientific.name2)

#loop through and update name to correct taxon
for(i in 1:nrow(bird_pred_df)) {
  sci_index = match(bird_pred_df$scientificName[i],avibase_df$Scientific.name,nomatch=0)
  if(sci_index==0){ # then is in IOC2.7 database
    sci_index = match(bird_pred_df$scientificName[i],avibase_ioc_df$Scientific.name,nomatch=0)
    if(sci_index!=0){ #if it matches, then we assign it to the correct taxonomy
      bird_pred_df$scientificName[i] = avibase_ioc_df$Scientific.name2[sci_index]
    }else{
      bird_pred_df$scientificName[i] = NA
    }
    
  }else{ # then is in the birdlife 3 database and assign it to the correct taxonomy
    bird_pred_df$scientificName[i] = avibase_df$Scientific.name2[sci_index]
  }
}

#saving those with missing taxonomy to a datafile
bird_pred_df %>%
  filter(scientificName == "") %>%
  write.csv("missingtaxonomy_birds.csv")

#removing those taxons with missing taxonomy
bird_pred_df2 = bird_pred_df %>%
  filter(scientificName != "")

#chopping off some columns
bird_pred_df2 = bird_pred_df2[,1:42]

redlist_birds_df$scientificName = as.character(redlist_birds_df$scientificName)

new_bird_df = left_join(x = bird_pred_df2, y = redlist_birds_df,  
                    by="scientificName")

clean_bird_df = read.csv("cleaned_bird_df.csv")

colnames(clean_bird_df)
colnames(new_bird_df)

new_bird_df = new_bird_df %>% 
  dplyr::select(-infraType, -infraName, -infraAuthority, 
         -redlistCriteria, -criteriaVersion)


new_clean_bird_df = rbind(clean_bird_df,new_bird_df)

write.csv(new_clean_bird_df,"cleaned_bird_df_updated_5_29.csv")

```

## Old Code

Old Code to match species
```{r format species name}

# fulldata_df=NULL
# redlist_df$MSW05_Binomial = redlist_df$scientificName
# amniote_df$MSW05_Binomial = paste(amniote_df$genus,amniote_df$species,sep = " ")
# pantheria93_df$MSW05_Binomial = as.character(pantheria93_df$MSW05_Binomial)
# 
# redlist_df$MSW05_Binomial = as.character(gsub("^([^\\s]*\\s[^\\s]*)\\s.*$", "\\1",redlist_df$scientificName)) 
# #https://stackoverflow.com/questions/7449564/regex-return-all-before-the-second-occurrence
# #returns everything before the second space, so that we get the binomial
# redlist_df$MSW05_Binomial[40:50]
# 
# fulldata_df = left_join(x = redlist_df,amniote_df,by="MSW05_Binomial")
# 
# fulldata_df = left_join(x = fulldata_df,pantheria93_df,by="MSW05_Binomial")
# 
# write.csv(fulldata_df,"fulldataset2.csv")

```

```{r check nas}

# fulldata_df = read.csv("fulldataset.csv",header = TRUE)
# 
# 
# # for (i in 1:ncol(fulldata_df)) {
# #   fulldata_df[,i][fulldata_df[,i]== "-999" | fulldata_df[,i]=="NA"] = NA
# # }
# 
# 
# for (i in 1:ncol(fulldata_df)) {
#   print(colnames(fulldata_df)[i])
#   print(table(is.na(fulldata_df[,i])))
# }
# 
# 

```
