---
title: "STAT 434 Project Modeling"
author: "Emma Barton and Ryan Vosbigian"
date: "5/21/2020"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

Types of models to use:

Classification models only (could convert classes to numeric 1 thru 4, but makes our prediction undesirable, plus not equal jumps between classes)

  1. Logistic Regression (pairwise/one-versus-one and one-versus-all)?
  2. Linear Discriminant Analysis
  3. Quadratic Discriminant Analysis
  4. K-Nearest Neighbor
  5. Classification Tree
  6. Bagged Trees
  7. Random Forests
  8. Boosted Forests
  9. Neural Networks
  10. Support Vector Machines
  
Additional Considerations for Modeling

  1. PCA of all diet and foraging variables
    a. All diet variables are highly correlated, because they each sum up to 100. Thus, using PCA could really simplify the models considerably
    b. Additionally, if PCA is performed on diet variables then the diet variables should not be scaled because they already are by nature of the data
  2. Changing sensitivity of classifications so it is harder to be classified as Least Concern
    a. ROC Curves for each model
  
# Set Up

```{r load packages}

library(tidyverse)
library(class)
library(tree)
library(randomForest)
library(MASS)

```

```{r set pathway}

## Local Pathways - copy/paste your local pathway to the github repository

# Emma
pathway = "/Users/emmabarton/Desktop/spring_2020/STAT434/project/"

# Ryan
#pathway = "~/GitHub/STAT-434---not-the-dodo-/"


```

```{r load data}

mammal_df = read.csv("cleaned_mammal_df.csv")
bird_df = read.csv("cleaned_bird_df_updated_5_29.csv")

```

```{r chopping off extra rows}

nrow(bird_df)

bird_df = bird_df[1:9640,]

```


# Preparing Data

## Transforming Variables

```{r changing variable types}

# Converting mammal activity variables to factors 
mammal_df$Activity.Nocturnal = as.factor(mammal_df$Activity.Nocturnal)
mammal_df$Activity.Crepuscular = as.factor(mammal_df$Activity.Crepuscular)
mammal_df$Activity.Diurnal = as.factor(mammal_df$Activity.Diurnal)
mammal_df$redlistCategory = as.factor(mammal_df$redlistCategory)

# Converting bird activity variables to factors 
bird_df$Nocturnal = as.factor(bird_df$Nocturnal)
bird_df$redlistCategory = as.factor(bird_df$redlistCategory)

```

```{r response variable transformations}

# Combine "Extinct" & "Extinct in the Wild" & "Critically Endangered" into "Endangered"
mammal_df$redlistCategory = as.character(car::recode(mammal_df$redlistCategory,recodes = "'Extinct'='Endangered';'Extinct in the Wild'='Endangered';'Critically Endangered'='Endangered'"))

bird_df$redlistCategory = as.character(car::recode(bird_df$redlistCategory,recodes = "'Extinct'='Endangered';'Extinct in the Wild'='Endangered';'Critically Endangered'='Endangered'"))

```

```{r explanatory variable transformations}

# Mammal Explanatory Variable Transformations
mammal_df = mammal_df %>% 
  mutate(
    log_bodymass = log(BodyMass.Value)
  )


# Bird Explanatory Variable Transformations 

bird_df = bird_df %>% 
  mutate(
    log_bodymass = log(BodyMass.Value)
  )
```

## Adding PCA of Diet and Foraging

Creating data frames with Principle Components as columns
```{r}

pcbirddiet = prcomp(bird_df[,c(12:21)],scale=FALSE)

pcbirdforage = prcomp(bird_df[,c(26:32)],scale=FALSE)

pcmammal = prcomp(mammal_df[,c(4:13)],scale = FALSE)

pcbird = prcomp(bird_df[,c(12:21,26:32)],scale=FALSE)

summary(pcmammal)


plot(cumsum(pcbirddiet$sdev^2/sum(pcbirddiet$sdev^2)), xlab="Principal  Component", ylab="Cumulative  Proportion  of Variance  Explained", ylim=c(0,1),type="b",main = "Bird Diet")

plot(cumsum(pcmammal$sdev^2/sum(pcmammal$sdev^2)), xlab="Principal  Component", ylab="Cumulative  Proportion  of Variance  Explained", ylim=c(0,1),type="b",main = "Mammal Diet")


plot(cumsum(pcbirdforage$sdev^2/sum(pcbirdforage$sdev^2)), xlab="Principal  Component", ylab="Cumulative  Proportion  of Variance  Explained", ylim=c(0,1),type="b",main = "Bird Foraging")

plot(cumsum(pcbird$sdev^2/sum(pcbird$sdev^2)), xlab="Principal  Component", ylab="Cumulative  Proportion  of Variance  Explained", ylim=c(0,1),type="b",main = "Bird Foraging and Diet")


colnames(pcbirddiet$x) = paste("Diet",colnames(pcbirddiet$x),sep = "")

colnames(pcmammal$x) = paste("Diet",colnames(pcmammal$x),sep = "")

colnames(pcbirdforage$x) = paste("For",colnames(pcbirdforage$x),sep = "")

colnames(pcbird$x) = paste("DietFor",colnames(pcbird$x),sep = "")

bird_df = cbind(bird_df,
                   cbind(pcbirddiet$x,
                         cbind(pcbirdforage$x,pcbird$x)))

mammal_df = cbind(mammal_df,pcmammal$x)

remove(pcbird,pcbirddiet,pcbirdforage,pcmammal)

```

## Making Training and Test Datasets

```{r separating data deficient observations}

# Moving data deficient mammals to different data set
data_deficient_mammals_df = mammal_df[mammal_df$redlistCategory == "Data Deficient",]
# Removing data deficient mammals
mammal_df = mammal_df[mammal_df$redlistCategory != "Data Deficient",]

# Moving data def birds to different dataset
data_deficient_birds_df = bird_df[bird_df$redlistCategory == "Data Deficient",]
# Removing data deficient birds
bird_df = bird_df[bird_df$redlistCategory != "Data Deficient",]

```

```{r creating test and training datasets}

# Creating indices for training sets
set.seed(10)
train_mammal = sample(1:nrow(mammal_df), size = round(nrow(mammal_df)/2))
train_bird = sample(1:nrow(bird_df), size = round(nrow(bird_df)/2))

# Creating training and test sets
mammal_train_df = mammal_df[train_mammal,]
mammal_test_df = mammal_df[-train_mammal,]

bird_train_df = bird_df[train_bird,]
bird_test_df = bird_df[-train_bird,]

```

# 1: Logistic Regression
  
# 2: Linear Discriminant Analysis

## Mammals

```{r}

library(MASS)



mam_lda1 = lda(redlistCategory ~ Diet.Vfish + Diet.Vend + Diet.Scav + Diet.Seed + Diet.Inv + Diet.Fruit + Diet.Nect +  + Diet.PlantO + ForStrat.Value + Activity.Nocturnal + Activity.Crepuscular +  Activity.Diurnal + log_bodymass, data=mammal_train_df)

mam_lda1_pred=predict(mam_lda1, mammal_test_df)

table(preds = mam_lda1_pred$class, truth = mammal_test_df$redlistCategory)

mean(mam_lda1_pred$class!=mammal_test_df$redlistCategory)
#error rate 0.3080149
mam_lda1

```

```{r}
mam_lda1 = lda(redlistCategory ~ Diet.Vfish + Diet.Vend + Diet.Scav + Diet.Seed + Diet.Inv + Diet.Fruit + Diet.Nect +  + Diet.PlantO + ForStrat.Value + Activity.Nocturnal + Activity.Crepuscular +  Activity.Diurnal + log_bodymass, data=mammal_train_df)

mam_lda1_pred=predict(mam_lda1, mammal_test_df)

table(preds = mam_lda1_pred$class, truth = mammal_test_df$redlistCategory)

mean(mam_lda1_pred$class!=mammal_test_df$redlistCategory)
#error rate 0.3080149
```


### Variable Selection

```{r}
mammal_lda_variables = list(c(38,4, 5, 6, 7, 8, 9, 10, 11, 12, 13), #all diet variables
                            c(38,7, 9, 10, 11, 12, 13),                  #selected diet variables
                            c(38,41),                               #log body mass
                            c(38,19, 20, 21, 41),                   #activity and body mass
                            c(38,4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 19, 20, 21), #all diet and activity variables
                            c(38,4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 41),         #all diet and body mass
                            c(38,41, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 19, 20, 21), #all diet and activity and body mass
                            c(38,7, 9, 10, 11, 12, 13, 41), #selected activity variables and body mass
                            c(38,7, 9, 10, 11, 12, 13, 19, 20, 21), #selected diet and activity
                            c(38,7, 9, 10, 11, 12, 13, 19, 20, 21, 41), #selected diet, activity and body mass
                            c(38,42, 43, 44, 45, 46, 47, 48), # 7 PC for diet
                            c(38,42, 43, 44, 45, 46, 47, 48, 41), # PCA for diet and diet
                            c(38,42, 43, 44, 45, 46, 47, 48, 19, 20, 21), # PCA for diet and activity
                            c(38,42, 43, 44, 45, 46, 47, 48, 19, 20, 21, 41), # PCA for diet, activity, and body mass
                            c(38,7, 9, 10, 11, 12, 13, 16),     #selected diet variables and foraging
                            c(38,7, 9, 10, 11, 12, 13, 41, 16), #selected activity variables and body mass and foraging
                            c(38, 16, 19, 20, 21), # activity and foraging
                            c(38, 16, 42, 43, 44, 45, 46, 47, 48, 41),   # PCA for diet, and body mass and foraging
                            c(38, 16, 42, 43, 44, 45, 46, 47, 48, 19, 20, 21, 41), #PCAdiet, and activity, body mass, and foraging
                            c(38,7, 9, 10, 11, 12, 13, 19, 20, 21, 41, 16)#selected diet, activity and body mass and foraging
                        
                            )  
test_errors = rep(0, length(mammal_lda_variables))

mammal_train_df[,mammal_lda_variables[[1]]]

for(i in 1:length(mammal_lda_variables)){
  var = mammal_lda_variables[[i]]
  mammal_lda = lda(redlistCategory ~ .,
                   data = mammal_train_df[,var])
  lda_pred = predict(mammal_lda,newdata=mammal_test_df[,var])
  test_errors[i] = mean(lda_pred$class != mammal_test_df$redlistCategory)
}

which.min(test_errors)
mammal_lda_variables[[which.min(test_errors)]]



```
### Best Model

```{r}

mammal_lda = lda(redlistCategory ~ .,
                   data = mammal_train_df[,mammal_lda_variables[[19]]])
lda_pred = predict(mammal_lda,newdata=mammal_test_df[,mammal_lda_variables[[19]]])
mean(lda_pred$class != mammal_test_df$redlistCategory)
table(pred=lda_pred$class,truth=mammal_test_df$redlistCategory)

#test error 0.3089469

colnames(mammal_train_df)[mammal_lda_variables[[19]]]
```



## Birds

```{r}

library(MASS)
library(pROC)

bird_lda1 = lda(redlistCategory ~ Diet.Inv + Diet.Vend + Diet.Vect + Diet.Vfish + Diet.Vunk + Diet.Scav + Diet.Fruit + Diet.Nect + Diet.Seed + Diet.PlantO + ForStrat.watbelowsurf + ForStrat.wataroundsurf + ForStrat.aerial +  PelagicSpecialist + Nocturnal + BodyMass.Value, data=bird_train_df)

bird_lda1_pred=predict(bird_lda1, bird_test_df)

table(preds = bird_lda1_pred$class, truth = bird_test_df$redlistCategory)

mean(bird_lda1_pred$class!=bird_test_df$redlistCategory)
#error rate 0.2315066


#Using first 11 PCs for Diet and Foraging PCs, because that is where the elbow is in the graph above
bird_lda_pca1 = lda(redlistCategory ~ DietForPC1 + DietForPC2 + DietForPC3 + DietForPC4 + DietForPC5 + DietForPC6 + DietForPC7 + DietForPC8 + DietForPC9 + DietForPC10 +DietForPC11 +  PelagicSpecialist + Nocturnal + scale(BodyMass.Value),data=bird_train_df)

bird_lda_pca1_pred=predict(bird_lda_pca1, bird_test_df)

table(preds = bird_lda_pca1_pred$class, truth = bird_test_df$redlistCategory)

mean(bird_lda_pca1_pred$class != bird_test_df$redlistCategory)
#error rate 0.2287977
colnames(bird_test_df)
```

### Variable Selection

```{r}

bird_lda_variables = list(c(54, 12:21), #all diet variables
                            c(54,91),                               #log body mass
                            c(54,37,91),                   #activity and body mass
                            c(54,12:21, 37), #all diet and activity variables
                            c(54,12:21, 91),         #all diet and body mass
                            c(54,12:21, 37, 91), #all diet and activity and body mass
                            c(54,26:33), # foraging variables
                            c(54,26:33, 12:21), # diet and foraging
                            c(54,26:33,91), #foraging and body mass
                            c(54,26:33, 12:21, 37, 91), # foraging, body mass, activity, and diet
                            c(54,109:119), # PCA for diet and foraing together
                            c(54,109:119, 91), # PCA for diet and foraging together and body mass
                            c(54,109:119, 91, 37) # PCA for diet + foraging, activity, and body mass
                            )
test_errors = rep(0, length(bird_lda_variables))



for(i in 1:length(bird_lda_variables)){
  var = bird_lda_variables[[i]]
  bird_lda = lda(redlistCategory ~ .,
                   data = bird_train_df[,var])
  lda_pred = predict(bird_lda,newdata=bird_test_df[,var])
  test_errors[i] = mean(lda_pred$class != bird_test_df$redlistCategory)
}

which.min(test_errors)
bird_lda_variables[[which.min(test_errors)]]
```


### Best Model

```{r}
bird_lda = lda(redlistCategory ~ .,
                   data = bird_train_df[,bird_lda_variables[[7]]])
lda_pred = predict(bird_lda,newdata=bird_test_df[,bird_lda_variables[[7]]])
mean(lda_pred$class != bird_test_df$redlistCategory)
table(pred=lda_pred$class,truth=bird_test_df$redlistCategory)

#test error 0.227339

colnames(bird_train_df)[mammal_lda_variables[[7]]]
```
This is the worst possible model.


# 3: Quadratic Discriminant Analysis

## Mammals

### Variable Selection

```{r}

# using less variables to avoid rank deficiency, just removing some that have collinearity
mammal_qda_variables = list(c(38,4, 5, 6, 7, 8, 9, 10, 11,13), #all diet variables
                            c(38,7, 9, 10, 11, 13),                  #selected diet variables
                            c(38,41),                               #log body mass
                            c(38,19, 20, 21, 41),                   #activity and body mass
                            c(38,4, 5, 6, 7, 8, 9, 10, 11, 19, 20), #all diet and activity variables
                            c(38,4, 5, 6, 7, 8, 9, 10, 11, 13, 41),         #all diet and body mass
                            c(38,41, 4, 5, 6, 7, 8, 9, 10, 11, 13, 19, 20, 21), #all diet and activity and body mass
                            c(38,7, 9, 10, 11, 13, 41), #selected activity variables and body mass
                            c(38,7, 9, 10, 11, 13, 19, 20, 21), #selected diet and activity
                            c(38,7, 9, 10, 11, 13, 19, 20, 21, 41), #selected diet, activity and body mass
                            c(38,42, 43, 44, 45, 46, 47, 48), # 7 PC for diet
                            c(38,42, 43, 44, 45, 46, 47, 48, 41), # PCA for diet and diet
                            c(38,42, 43, 44, 45, 46, 47, 48, 19, 20, 21), # PCA for diet and activity
                            c(38,42, 43, 44, 45, 46, 47, 48, 19, 20, 21, 41), # PCA for diet, activity, and body mass
                            c(38,7, 9, 10, 11, 12, 13, 16),     #selected diet variables and foraging
                            c(38,7, 9, 10, 11, 12, 13, 41, 16), #selected activity variables and body mass and foraging
                            c(38, 16, 19, 20, 21), # activity and foraging
                            c(38, 16, 42, 43, 44, 45, 46, 47, 48, 41),   # PCA for diet, and body mass and foraging
                            c(38, 16, 42, 43, 44, 45, 46, 47, 48, 19, 20, 21, 41), #PCAdiet, and activity, body mass, and foraging
                            c(38,7, 9, 10, 11, 12, 13, 19, 20, 21, 41, 16)#selected diet, activity and body mass and foraging
                        
                            )  
test_errors = rep(0, length(mammal_lda_variables))



for(i in 1:length(mammal_qda_variables)){
  var = mammal_qda_variables[[i]]
  mammal_qda = qda(redlistCategory ~ .,
                   data = mammal_train_df[,var])
  qda_pred = predict(mammal_qda,newdata=mammal_test_df[,var])
  test_errors[i] = mean(qda_pred$class != mammal_test_df$redlistCategory)
}

which.min(test_errors)
mammal_qda_variables[[which.min(test_errors)]]
```

### Best Model

```{r}
mammal_qda = qda(redlistCategory ~ .,
                   data = mammal_train_df[,mammal_qda_variables[[3]]])
qda_pred = predict(mammal_qda,newdata=mammal_test_df[,mammal_qda_variables[[3]]])
mean(qda_pred$class != mammal_test_df$redlistCategory)
table(pred=qda_pred$class,truth=mammal_test_df$redlistCategory)

#test error 0.3289842

colnames(mammal_train_df)[mammal_qda_variables[[3]]]
```


## Birds

### Variable Selection

```{r}
#removing first diet and foraging variables to avoid rank deficiency
bird_qda_variables = list(c(54, 13:21), #all diet variables
                            c(54,91),                               #log body mass
                            c(54,37,91),                   #activity and body mass
                            c(54,13:21, 37), #all diet and activity variables
                            c(54,13:21, 91),         #all diet and body mass
                            c(54,13:21, 37, 91), #all diet and activity and body mass
                            c(54,27:33), # foraging variables
                            c(54,27:33, 13:21), # diet and foraging
                            c(54,27:33,91), #foraging and body mass
                            c(54,27:33, 13:21, 37, 91), # foraging, body mass, activity, and diet
                            c(54,109:119), # PCA for diet and foraing together
                            c(54,109:119, 91), # PCA for diet and foraging together and body mass
                            c(54,109:119, 91, 37) # PCA for diet + foraging, activity, and body mass
                            )
test_errors = rep(0, length(bird_qda_variables))



for(i in 1:length(bird_qda_variables)){
  var = bird_qda_variables[[i]]
  bird_qda = qda(redlistCategory ~ .,
                   data = bird_train_df[,var])
  qda_pred = predict(bird_qda,newdata=bird_test_df[,var])
  test_errors[i] = mean(qda_pred$class != bird_test_df$redlistCategory)
}

which.min(test_errors)
bird_qda_variables[[which.min(test_errors)]]

```


### Best Model

```{r}
bird_qda = qda(redlistCategory ~ .,
                   data = bird_train_df[,bird_qda_variables[[2]]])
qda_pred = predict(bird_qda,newdata=bird_test_df[,bird_qda_variables[[2]]])
mean(qda_pred$class != bird_test_df$redlistCategory)
table(pred=qda_pred$class,truth=bird_test_df$redlistCategory)

#test error 0.2279642

colnames(bird_train_df)[bird_qda_variables[[2]]]
```


# 4: K-Nearest Neighbor

## Mammal K-Nearest Neighbor

### Format Variables
For some reason the knn function cannot deal with explanatory variables as factors so we'll have to recode categorical variables such as ForStrat.Value
```{r mammal change factors to numeric}

# Create separate dataframes for KNN
mammal_knn_train_df = mammal_train_df
mammal_knn_test_df = mammal_test_df

# Change factor variables to numeric
mammal_knn_train_df = mammal_knn_train_df %>% 
  mutate(
    Activity.Diurnal = as.numeric(Activity.Diurnal),
    Activity.Crepuscular = as.numeric(Activity.Crepuscular),
    Activity.Nocturnal = as.numeric(Activity.Nocturnal),
    
    for_strat_num = case_when(
      ForStrat.Value == "A" ~ 1,
      ForStrat.Value == "Ar" ~ 2,
      ForStrat.Value == "G" ~ 3,
      ForStrat.Value == "M" ~ 4,
      ForStrat.Value == "S" ~ 5
      ),
    for_strat_num = as.numeric(for_strat_num),
    
    pop_trend_num = case_when(
      populationTrend == "Decreasing" ~ 1,
      populationTrend == "Increasing" ~ 2,
      populationTrend == "Stable" ~ 3,
      populationTrend == "Unknown" ~ 4   
      ),
    pop_trend_num = as.numeric(pop_trend_num)
  )

mammal_knn_test_df = mammal_knn_test_df %>% 
  mutate(
    Activity.Diurnal = as.numeric(Activity.Diurnal),
    Activity.Crepuscular = as.numeric(Activity.Crepuscular),
    Activity.Nocturnal = as.numeric(Activity.Nocturnal),
    
    for_strat_num = case_when(
      ForStrat.Value == "A" ~ 1,
      ForStrat.Value == "Ar" ~ 2,
      ForStrat.Value == "G" ~ 3,
      ForStrat.Value == "M" ~ 4,
      ForStrat.Value == "S" ~ 5
      ),
    for_strat_num = as.numeric(for_strat_num),
    
    pop_trend_num = case_when(
      populationTrend == "Decreasing" ~ 1,
      populationTrend == "Increasing" ~ 2,
      populationTrend == "Stable" ~ 3,
      populationTrend == "Unknown" ~ 4   
      ),
    pop_trend_num = as.numeric(pop_trend_num)
  )
#colnames(mammal_knn_train_df)
```

### Variable Selection
Explanatory Variables: Diet, Foraging Stratum, Activity, Body Mass, Population Trend. It appears that you cannot make single explanatory variable KNN models.
```{r mammal knn variable selection}

### Hey Emma! Can you use these sets of variables? It also allows us to use the PCA

# list(c(38,4, 5, 6, 7, 8, 9, 10, 11, 12, 13), #all diet variables
#                             c(38,7, 9, 10, 11, 12, 13),                  #selected diet variables
#                             c(38,41),                               #log body mass
#                             c(38,19, 20, 21, 41),                   #activity and body mass
#                             c(38,4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 19, 20, 21), #all diet and activity variables
#                             c(38,4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 41),         #all diet and body mass
#                             c(38,41, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 19, 20, 21), #all diet and activity and body mass
#                             c(38,7, 9, 10, 11, 12, 13, 41), #selected activity variables and body mass
#                             c(38,7, 9, 10, 11, 12, 13, 19, 20, 21), #selected diet and activity
#                             c(38,7, 9, 10, 11, 12, 13, 19, 20, 21, 41), #selected diet, activity and body mass
#                             c(38,42, 43, 44, 45, 46, 47, 48), # 7 PC for diet
#                             c(38,42, 43, 44, 45, 46, 47, 48, 41), # PCA for diet and diet
#                             c(38,42, 43, 44, 45, 46, 47, 48, 19, 20, 21), # PCA for diet and activity
#                             c(38,42, 43, 44, 45, 46, 47, 48, 19, 20, 21, 41), # PCA for diet, activity, and body mass
#                             c(38,7, 9, 10, 11, 12, 13, 16),     #selected diet variables and foraging
#                             c(38,7, 9, 10, 11, 12, 13, 41, 16), #selected activity variables and body mass and foraging
#                             c(38, 16, 19, 20, 21), # activity and foraging
#                             c(38, 16, 42, 43, 44, 45, 46, 47, 48, 41),   # PCA for diet, and body mass and foraging
#                             c(38, 16, 42, 43, 44, 45, 46, 47, 48, 19, 20, 21, 41), #PCAdiet, and activity, body mass, and foraging
#                             c(38,7, 9, 10, 11, 12, 13, 19, 20, 21, 41, 16)#selected diet, activity and body mass and foraging
#                         
#                             )  


mammal_knn_variables = list(c(4, 5, 6, 7, 8, 9, 10, 11, 12, 13), 
                            c(41, 43),
                            c(41, 42, 43),
                            c(19, 20, 21, 41, 42, 43),
                            c(4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 19, 20, 21),
                            c(4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 41, 42, 43),
                            c(41, 42, 43, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 19, 20, 21)) 
test_errors = rep(0, length(mammal_knn_variables))

for(i in 1:length(mammal_knn_variables)){
  var = mammal_knn_variables[[i]]
  mammal_knn = knn(mammal_knn_train_df[,var],
                 mammal_knn_test_df[,var],
                 mammal_knn_train_df[,38], k=14)
  test_errors[i] = mean(mammal_knn != mammal_knn_test_df$redlistCategory)
}

which.min(test_errors)
mammal_knn_variables[[which.min(test_errors)]]

```

```{r mammal prediction and error rate}

# Error Rate
mean(mammal_knn != mammal_knn_test_df$redlistCategory)

table(preds = mammal_knn, truth = mammal_knn_test_df$redlistCategory)

```

### Hyperparameter Tuning

```{r mammal k value selection}

test_errors = rep(0, 25)
for(i in 1:25){
  mammal_knn = knn(mammal_knn_train_df[,c(4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 41, 42, 43)],
                 mammal_knn_test_df[,c(4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 41, 42, 43)],
                 mammal_knn_train_df[,38], k=i)
  test_errors[i] = mean(mammal_knn != mammal_knn_test_df$redlistCategory)
}

which.min(test_errors)

```

### Best Model

```{r mammal best knn model}

mammal_knn = knn(mammal_knn_train_df[,c(4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 41, 42, 43)],
                 mammal_knn_test_df[,c(4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 41, 42, 43)],
                 mammal_knn_train_df[,38], k=14)
mean(mammal_knn != mammal_knn_test_df$redlistCategory)

#error rate 0.3000932
```

## Bird K-Nearest Neighbor

### Format Variables

```{r bird change factors to numeric}

bird_knn_train_df = bird_train_df
bird_knn_test_df = bird_test_df

```

### Variable Selection

### Hyperparameter Tuning

### Best Model

# 5: Classification Tree

## Mammal Classification Tree

```{r mammal change variable types}

mammal_train_df$redlistCategory = as.factor(mammal_train_df$redlistCategory)
mammal_test_df$redlistCategory = as.factor(mammal_test_df$redlistCategory)

```

```{r mammal class tree}

mammal_tree = tree(redlistCategory ~ log_bodymass + populationTrend + ForStrat.Value + 
                   Activity.Crepuscular + Activity.Diurnal + Activity.Nocturnal + 
                   Diet.Inv + Diet.Vend + Diet.Vect + Diet.Vfish + Diet.Vunk + 
                   Diet.Scav + Diet.Fruit + Diet.PlantO, 
                   mammal_train_df)

```

```{r mammal tree assessment}

summary(mammal_tree)

plot(mammal_tree)
text(mammal_tree, pretty = 0)

```


# 6: Bagged Trees

## Mammal Bagged Trees

### Variable Selection

```{r}
set.seed(10)
bag.mam1 = randomForest(as.factor(redlistCategory) ~ Diet.Vfish + Diet.Vend + Diet.Scav + Diet.Seed + Diet.Inv + 
                          Diet.Fruit + Diet.Nect +  + Diet.PlantO + ForStrat.Value + Activity.Nocturnal +
                          Activity.Crepuscular +  Activity.Diurnal + log_bodymass,
                        data=mammal_train_df,
                        mtry=13,
                        importance=TRUE)

bag.mam1

yhat.bag = predict(bag.mam1 ,newdata=mammal_test_df)

table(yhat.bag,mammal_knn_test_df$redlistCategory)

mean(yhat.bag!=mammal_knn_test_df$redlistCategory)
#test error 0.3657968

bag.mam1$importance

```

Dropping leaset important variable, based on Gini index: Diet.Vfish

```{r}
set.seed(10)
bag.mam2 = randomForest(as.factor(redlistCategory) ~ Diet.Vend + Diet.Scav + Diet.Seed + Diet.Inv + 
                          Diet.Fruit + Diet.Nect +  + Diet.PlantO + ForStrat.Value + Activity.Nocturnal +
                          Activity.Crepuscular +  Activity.Diurnal + log_bodymass,
                        data=mammal_train_df,
                        mtry=12,
                        importance=TRUE)

bag.mam2

yhat.bag = predict(bag.mam2 ,newdata=mammal_test_df)

table(yhat.bag,mammal_knn_test_df$redlistCategory)

mean(yhat.bag!=mammal_knn_test_df$redlistCategory)
#test error 0.3648649

bag.mam2$importance
```
# Now dropping: Diet.Scav

```{r}
set.seed(10)
bag.mam3 = randomForest(as.factor(redlistCategory) ~ Diet.Vend + Diet.Seed + Diet.Inv + 
                          Diet.Fruit + Diet.Nect +  + Diet.PlantO + ForStrat.Value + Activity.Nocturnal +
                          Activity.Crepuscular +  Activity.Diurnal + log_bodymass,
                        data=mammal_train_df,
                        mtry=11,
                        importance=TRUE)

bag.mam3

yhat.bag = predict(bag.mam3 ,newdata=mammal_test_df)

table(yhat.bag,mammal_knn_test_df$redlistCategory)

mean(yhat.bag!=mammal_knn_test_df$redlistCategory)
#test error 0.3667288

bag.mam3$importance
```


#### Using PCA

```{r}

set.seed(10)
bag.mam.pca1 = randomForest(as.factor(redlistCategory) ~ DietPC1 + DietPC2 + DietPC3 + DietPC4 + DietPC5 + 
                          DietPC6 + DietPC7 + ForStrat.Value + Activity.Nocturnal +
                          Activity.Crepuscular +  Activity.Diurnal + log_bodymass,
                        data=mammal_train_df,
                        mtry=12,
                        importance=TRUE)

bag.mam.pca1

yhat.bag = predict(bag.mam.pca1 ,newdata=mammal_test_df)

table(yhat.bag,mammal_test_df$redlistCategory)

mean(yhat.bag!=mammal_test_df$redlistCategory)
# test error 0.3699907

bag.mam.pca1$importance
```

### Best Model

```{r}
set.seed(10)
bag.mam1 = randomForest(as.factor(redlistCategory) ~ Diet.Vfish + Diet.Vend + Diet.Scav + Diet.Seed + Diet.Inv + 
                          Diet.Fruit + Diet.Nect +  + Diet.PlantO + ForStrat.Value + Activity.Nocturnal +
                          Activity.Crepuscular +  Activity.Diurnal + log_bodymass,
                        data=mammal_train_df,
                        mtry=13,
                        importance=TRUE)

bag.mam1

yhat.bag = predict(bag.mam1 ,newdata=mammal_test_df)

table(yhat.bag,mammal_knn_test_df$redlistCategory)

mean(yhat.bag!=mammal_knn_test_df$redlistCategory)
#test error 0.3657968

```


## Bird Bagged Trees

### Variable Selection

```{r}
set.seed(10)
bag.bird1 = randomForest(as.factor(redlistCategory) ~ Diet.Inv + Diet.Vend + Diet.Vect + Diet.Vfish + Diet.Vunk + 
                          Diet.Scav + Diet.Fruit + Diet.Nect + Diet.Seed + Diet.PlantO +  
                          ForStrat.watbelowsurf + ForStrat.wataroundsurf + ForStrat.aerial +  
                          PelagicSpecialist + Nocturnal + log_bodymass,
                        data=bird_train_df,
                        mtry=16,
                        importance=TRUE)

bag.bird1

yhat.bag = predict(bag.bird1 ,newdata=bird_test_df)

table(yhat.bag,bird_test_df$redlistCategory)

mean(yhat.bag!=bird_test_df$redlistCategory)
#test error 0.2821421

bag.bird1$importance

```
#### Using Principle Components

```{r}
set.seed(10)
bag.bird2 = randomForest(as.factor(redlistCategory) ~ DietForPC1 + DietForPC2 + DietForPC3 + DietForPC4 + 
                           DietForPC5 + DietForPC6 + DietForPC7 + DietForPC8 + DietForPC9 + DietForPC10 +DietForPC11 +  
                          Nocturnal + log_bodymass,
                        data=bird_train_df,
                        mtry=13,
                        importance=TRUE)

bag.bird2

yhat.bag = predict(bag.bird2 ,newdata=bird_test_df)

table(yhat.bag,bird_test_df$redlistCategory)

mean(yhat.bag!=bird_test_df$redlistCategory)
#test error 0.2783913

bag.bird2$importance
```


### Best Model

```{r}
set.seed(10)
bag.bird2 = randomForest(as.factor(redlistCategory) ~ DietForPC1 + DietForPC2 + DietForPC3 + DietForPC4 + 
                           DietForPC5 + DietForPC6 + DietForPC7 + DietForPC8 + DietForPC9 + DietForPC10 +DietForPC11 +  
                          Nocturnal + log_bodymass,
                        data=bird_train_df,
                        mtry=13,
                        importance=TRUE)

bag.bird2

yhat.bag = predict(bag.bird2 ,newdata=bird_test_df)

table(yhat.bag,bird_test_df$redlistCategory)

mean(yhat.bag!=bird_test_df$redlistCategory)
#test error 0.2783913

```


# 7: Random Forests

# 8: Boosted Forests



# 9: Neural Networks

# 10: Support Vector Machines








